D3064R0: How to Avoid OOTA Without Really Trying

### 2024-02-23 ootavacuous.tex
Update the commentary on the causality test cases
The commentary on the causality test cases is out of date as compared
with the rest of the paper.  Update it, while removing some unneeded
parts.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Listing~\ref{lst:Causality Test Case 1}
shows causality test case 1, for which the \co{r1 == r2 == 1} result
is to be allowed.
The \co{herd7} tool finds this litmus test's cycle, even without the
inter-thread analysis, because C++ does not respect control dependencies
to relaxed atomic stores.

@@

Implementations that can prove that the value of \co{x} is always
nonnegative can also prove that there is no semantic dependency between
\co{P0()}'s load from \co{x} on line~8 to its store to \co{y} on line~10,
that is, no matter which of the possible values is loaded from \co{x},
the value 1 will always be stored to \co{y}.

Less-omniscient implementations, including those whose code analysis
is restricted to a thread, would need to act as if there was in fact a
semantic dependency.

Even omniscient implementations will observe a semantic dependency from
\co{P1()}'s load from \co{y} on line~14 to its store to \co{x} on line~15.

Either way, there is no OOTA cycle, with omniscient implementations
having no cycle at all, and other implementations retaining the
nonsemantic dependency.

@@

Listing~\ref{lst:Causality Test Case 2}
shows causality test case 2, for which the \co{r1 == r2 == r3 == 1} result
is to be allowed.
The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.

@@

Because \co{P0()}'s loads from \co{x} on lines~8 and~9 are unordered,
an implementation can act as if the source code loaded from \co{x} only
once and assigned the returned value to both \co{r1} and \co{r2}.\footnote{
	But not in C, where relaxed atomic operations are volatile,
	and thus observable behavior.}
The implementation could then prove that the condition on line~10
would always evaluate to \co{true}, and act as if the source code had
unconditionally executed the relaxed store to \co{y} on line~11.
The implementation could then act as if the resulting relaxed accesses
to \co{x} and \co{y} appeared in reverse order, at which point
\co{P0()}'s store to \co{y} could be read by \co{P1()} load so that
\co{P1()} stored 1 to \co{x}, thus resulting in \co{P1()} loading
that 1 from \co{x}.

Note that \co{P1()} has a trivial semantic dependency from its load
from \co{y} on line~15 to its store to \co{x} on line~16.

If \co{P0()}'s loads from \co{x} are fused, there is no cycle,
otherwise the preserved nonsemantic dependency prevents an
OOTA cycle from forming on correct real-world implementations.

@@

Listing~\ref{lst:Causality Test Case 3}
shows causality test case 3, for which the \co{r1 == r2 == r3 == 1} result
is to be allowed.
The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.

The analysis from Section~\ref{app:Causality Test Case 2}
applies here, including the difference in volatility between C and C++
atomics.
Although the addition of \co{P2} allows sequentially consistent executions
where \co{P0()}'s two loads from \co{x} might return different values,
it does not change the fact that those two loads can be merged.

@@

Listing~\ref{lst:Causality Test Case 4}
shows causality test case 4, for which the \co{r1 == r2 == 1} result
is to be forbidden.
The \co{herd7} tool finds this OOTA cycle because C++ does not respect
data dependencies, and flags it as such using \co{S8}.

@@

Both \co{P0()} and \co{P1()} have straightforward semantic dependencies
from their relaxed atomic loads to their relaxed atomic stores, which
prevents this OOTA cycle on correct real-world C++ implementations.

@@

Listing~\ref{lst:Causality Test Case 5}
shows causality test case 5, for which the \co{r1 == r2 == 1 && r3 == 0}
result is to be forbidden.
The \co{herd7} tool finds this OOTA cycle because C++ does not respect
data dependencies, and flags it as such using \co{S8}.

@@

Each of \co{P0()}, \co{P1()}, and \co{P3()} have trivial semantic
dependencies from their relaxed atomic loads to their relaxed
atomic stores, which prevents any OOTA cycle from forming in correct
real-world C++ implementations.
\co{P2()} has no loads, so cannot participate in an OOTA cycle in
any case.

@@

Listing~\ref{lst:Causality Test Case 6}
shows causality test case 6, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool finds this cycle, and does not flag it as an OOTA
cycle because C++ does not respect control dependencies to relaxed
atomic stores.

@@

A C++ implementation that can prove that \co{y} is always either
zero or one can also prove that there is no semantic dependency
from \co{P1()}'s load on line~14 to its stores on lines~16 and~29.
Such an implementation could act as if lines~15-20 had been replaced
by a single store to \co{x}, allowing that store to execute first,
followed by lines~8, 10, and~14, in that order, satisfying the
\co{exists} clause on line~23.

Less-omniscient implementations will preserve the nonsemantic
dependency, preventing any OOTA loop from forming in a correct real-world C++
implementation.

@@

Listing~\ref{lst:Causality Test Case 7}
shows causality test case 7, for which the \co{r1 == r2 == r3 == 1}
result is to be allowed.
The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.

@@

Because all accesses are relaxed atomics, this test case has everything
to do with reordering and nothing to do with OOTA cycles.
The outcome happens when lines~17, 10, 11, 15, 16 and~9 execute in that
order.
Note that this ordering preserves the semantic dependencies from
lines~10-11 and~15-16, and can thus happen on correct real-world
C++ implementations.

@@

Listing~\ref{lst:Causality Test Case 8}
shows causality test case 8, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
initialized to the constant \co{1}, it does find the cycle and refrains
from flagging it as an OOTA cycle.

@@

Omniscient C++ implementations might note that the computation on line~9
can only have the value one when the load from \co{x} returns either
zero or one, and both \co{x} and \co{y} are initialized to zero.
Such an implementation act as if line~10's \co{r2} had instead been
the value~1, so that the value returned by \co{P0()}'s load could not
affect the value stored.
Then there would be no semantic dependency between \co{P0()}'s load
and store, and thus no OOTA cycle in a correct real-world C++
implementation.

A less-capable implementation would leave the nonsemantic dependency
in place, also preventing any OOTA cycle in a correct real-world C++
implementation.

@@

Listing~\ref{lst:Causality Test Case 9}
shows causality test case 9, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not find this cycle because it depends on
scheduling \co{P2()} after the other two threads, and C++ is not
constrained to do this.

@@

Both threads have straightforward semantic dependencies between their
respective relaxed atomic loads and relaxed atomic stores, so an OOTA
cycle cannot form in a correct real-world C++ implementation.

@@

Listing~\ref{lst:Causality Test Case 9a}
shows causality test case 9a, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not find this cycle because it depends on
scheduling \co{P2()} before the other two threads, and C++ is not
constrained to do this.

@@

Both threads have straightforward semantic dependencies between their
respective relaxed atomic loads and relaxed atomic stores, so an OOTA
cycle cannot form in a correct real-world C++ implementation.

@@

Listing~\ref{lst:Causality Test Case 10}
shows causality test case 10, for which the \co{r1 == r2 == 1 && r3 == 0}
result is to be forbidden.
Because C++ does not respect control dependencies, the \co{herd7} tool
nevertheless reports this result and does not flag it as an OOTA cycle.

@@

However, each of \co{P0()}, \co{P1()}, and \co{P3()} have straightforward
semantic dependencies between their respective loads and stores, and
thus cannot form an OOTA cycle when running on a correct real-world C++
implementation.
For its part, \co{P2()} has no semantic dependency, so it cannot be part
of an OOTA cycle when running on a correct real-world C++ implementation.

@@

Listing~\ref{lst:Causality Test Case 11}
shows causality test case 11, for which the \co{r1 == r2 == r3 == r4 == 1}
result is to be allowed.
The \co{herd7} tool reports this result and does not flag it as an OOTA cycle.

@@

This test case has nothing to do with OOTA cycles.
The result may be obtained by executing the statements in the following
order: line~20, line~12, line~13, line~18, line~19, line~10, line~11,
and finally line~17.

@@

JMM Causality Test Case 12 in
Listing~\ref{lst:Causality Test Case 12}
uses arrays, which are not yet supported by the \co{herd7} tool.

@@

An omniscient C++ implementation could note that only \co{P0()} accesses
array \co{a[]} and act as if the code was simpler, but there would
still be semantic dependencies between \co{P0()}'s and \co{P1()}s
relaxed atomic loads and stores involving \co{x} and \co{y}.
Those dependencies prevent an OOTA cycle from forming on a correct
real-world C++ implementation, omniscient or not.

@@

Listing~\ref{lst:Causality Test Case 13}
shows causality test case 13, for which the \co{r1 == r2 == 1}
result is to be forbidden.
But because C++ does not respect control dependencies, \co{herd7} tool
reports this result and does not flag it as an OOTA cycle.

@@

There are straightforward semantic dependencies between \co{P0()}'s and
\co{P1()}s relaxed atomic loads and stores involving \co{x} and \co{y}.
Those dependencies prevent an OOTA cycle from forming on a correct
real-world C++ implementation.

@@

Listing~\ref{lst:Causality Test Case 14}
shows causality test case 14, for which the \co{r1 == r3 == 1 && r2 == 0}
result is to be forbidden.
The \co{herd7} tool nevertheless reports this result because C++ does not
respect control dependencies for atomic stores.

@@

In C++, this is not a matter of OOTA but rather reordering.
Dependencies notwithstanding, according to the C++ memory model,
the statements could execute in the following order:
Line~17 (so \co{r2 == 0}), line~21, line~9 (so \co{r1 == 1}),
line~13, line~18 (so \co{r3 == 1)}.
The sum of \co{r2} and \co{r3} is nonzero, justifying line~21's
store to \co{x}.

However, correct real-world C++ implementations would have semantic
dependencies between \co{P0()}'s line~9 and lines~11 and~13, and also
between \co{P1()}'s lines~17 and~18 and line~21, preventing any
OOTA cycle from forming.

@@

Listing~\ref{lst:Causality Test Case 15}
shows causality test case 15, for which the \co{r0 == r1 == r3 == 1 && r2 == 0}
result is to be forbidden.
The \co{herd7} tool nevertheless reports this result because C++ does not
respect control dependencies for atomic stores.

@@

In C++, this is not a matter of OOTA but rather reordering.
Dependencies notwithstanding, according to the C++ memory model,
the statements could execute in the following order:
Line~31, line~27, line~10 (so \co{r0 == 1}), line~13 (so \co{r1 == 1}),
line~19 (because \co{r1 == 1}), line~23 (so \co{r2 == 0}),
and finally line~24 (so \co{r3 == 1}).
The sum of \co{r2} and \co{r3} is nonzero, justifying line~27's
store to \co{u}.

However, correct real-world C++ implementations would have semantic
dependencies between \co{P0()}'s lines~10 and~13 and lines~17 and~19, and also
between \co{P1()}'s lines~23 and~24 and line~27, preventing any
OOTA cycle from forming.
Having no load, \co{P2()} cannot participate in an OOTA cycle.

@@

Listing~\ref{lst:Causality Test Case 16}
shows causality test case 16, for which the \co{r1 == 2 && r2 == 1}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given that C++ does not permit a thread to reorder accesses to a single
atomic variable, even when those accesses are relaxed.

@@

Listing~\ref{lst:Causality Test Case 17}
shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given that C++ does not permit a thread to reorder accesses to a single
atomic variable, even when those accesses are relaxed.
But note that the \co{herd7} tool correctly notes the presence of an
OOTA cycle with \co{S17}.

@@

\co{P1()} has a straightforward semantic dependency between its
relaxed atomic load and relaxed atomic store.
However, suppose that \co{P0()}'s load from \co{x} on line~8 returns
the value 42.
In that case, the second load on line~11 can be fused with the first,
so that it also returns the value 42.
On the other hand, if the load on line~8 returns some other value, then
line~10 will store the value 42 to \co{x}, and then the load on line~11
can be elided\footnote{
	In C++, but not in C due to atomics being volatile.}
in favor of using the value 42 stored by line~10.

Either way, the value of \co{r1} will be the constant 42.
Therefore, the loads from \co{x} cannot affect the store on line~17,
so there is no semantic dependency.
This allows that store to \co{y} to be moved to the very beginning of
\co{P0()}, thus allowing a correct real-world C++ implementation to
satisfy the \co{exists} clause.

@@

Listing~\ref{lst:Causality Test Case 18}
shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given that C++ does not permit a thread to reorder accesses to a single
atomic variable, even when those accesses are relaxed.
But note that the \co{herd7} tool correctly notes the presence of an
OOTA cycle with \co{S17}.

@@

Both threads have straightforward semantic dependencies between their
respective relaxed atomic loads and relaxed atomic stores, so an OOTA
cycle cannot form in a correct real-world C++ implementation.

@@

Listing~\ref{lst:Causality Test Case 19}
shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given that C++ does not permit a thread to reorder accesses to a single
atomic variable, even when those accesses are relaxed.
But note that the \co{herd7} tool correctly notes the presence of an
OOTA cycle with \co{S8}.

It is interesting that this \co{S8} value does not propagate to
\co{r3}.

@@

All threads have straightforward semantic dependencies between their
respective relaxed atomic loads and relaxed atomic stores, so an OOTA
cycle cannot form in a correct real-world C++ implementation.

@@

Listing~\ref{lst:Causality Test Case 20}
shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be expected
given that C++ does not permit a thread to reorder accesses to a single
atomic variable, even when those accesses are relaxed.
But note that the \co{herd7} tool correctly notes the presence of an
OOTA cycle with \co{S8}.

It is interesting that this \co{S8} value does not propagate to
\co{r3}.

@@

All threads have straightforward semantic dependencies between their
respective relaxed atomic loads and relaxed atomic stores, so an OOTA
cycle cannot form in a correct real-world C++ implementation.

>>>>

Listing~\ref{lst:Causality Test Case 1}
shows causality test case 1, for which the \co{r1 == r2 == 1} result
is to be allowed.
And indeed, this result does show up in the output from the \co{herd7} tool.

@@

Compilers that can prove the value of \co{x} is always
nonnegative can also determine that there is no semantic dependency between
\co{P0()}'s load from \co{x} on line~8 and its store to \co{y} on line~10,
that is, no matter which of the possible values is loaded from \co{x},
the value 1 will always be stored to \co{y}.
Relative to less-omniscient compilers, however,
including those using single-thread analysis or treating atomic
objects as volatile, there \emph{is} a semantic dependency.

Even omniscient compilers will observe a semantic dependency from
\co{P1()}'s load from \co{y} on line~14 to its store to \co{x} on line~15.

Either way no OOTA cycle will occur, with omniscient compilers
having no cycle at all because of the lack of an sdep link in \co{P0()},
and other implementations prevented by hardware
dependency ordering from realizing the cycle.

@@

Listing~\ref{lst:Causality Test Case 2}
shows causality test case 2, for which the \co{r1 == r2 == r3 == 1} result
is to be allowed.
This result appears in the \co{herd7} output.

@@

Because \co{P0()}'s loads from \co{x} on lines~8 and~9 are unordered,
a C++ compiler that treats atomics as quasi volatile
can act as if the source code loaded from \co{x} only
once and assigned the returned value to both \co{r1} and \co{r2},
merging the loads.\footnote{
	But not in C, where relaxed atomic operations are volatile,
	and thus are observable behavior not subject to merging.}
The compiler could then prove that the condition on line~10
would always true, and act as if the source code had
unconditionally executed the relaxed store to \co{y} on line~11
independent of the value loaded from \co{x}.
Without this dependency there would then be no OOTA cycle.

For compilers that do not merge the two loads, the hardware
dependency ordering arising from the conditional test on line~10
will prevent the realization of an OOTA cycle.

@@

Listing~\ref{lst:Causality Test Case 3}
shows causality test case 3, for which the \co{r1 == r2 == r3 == 1} result
is to be allowed.
The \co{herd7} tool reports this result.

The analysis from the previous section applies here.
Although the addition of \co{P2()} allows sequentially consistent executions
in which \co{P0()}'s two loads from \co{x} return different values,
it does not change the fact that those two loads can be merged.

@@

Listing~\ref{lst:Causality Test Case 4}
shows causality test case 4, for which the \co{r1 == r2 == 1} result
is to be forbidden.
The \co{herd7} tool finds this OOTA result;
the \co{S8} values given on line~3 of its output indicate
that in this execution the variables \co{r1} and~\co{r2} were never
assigned any specific value, so they could just as well be equal to
one as to anything else.

@@

This test case is virtually the same as the Simple OOTA cycle
discussed in section~\ref{sec:Simple OOTA Cycle}, the only difference
being the final values of the variables.

@@

Listing~\ref{lst:Causality Test Case 5}
shows causality test case 5, for which the \co{r1 == r2 == 1} and \co{r3 == 0}
result is to be forbidden.
The \co{herd7} tool reports this OOTA result in line~3 of its output.

@@

This is the same as causality test case~4 with \co{P2()} and \co{P3()}
added, but those routines play no role in the OOTA cycle.
\co{P2()} has no loads and so cannot participate in an OOTA cycle in
any case.

@@

Listing~\ref{lst:Causality Test Case 6}
shows causality test case 6, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool reports this result.

@@

A compiler that can prove \co{y} is always either
zero or one can also prove that there is no semantic dependency
from \co{P1()}'s load on line~14 to its stores on lines~16 and~19.
Such a compiler could act as if lines~15--20 had been replaced
by a single store to \co{x}, showing that the dependency isn't
semantic.

Less-omniscient compilers will preserve the dependency, and then
hardware dependency ordering will prevent the OOTA cycle from
being realized.

@@

Listing~\ref{lst:Causality Test Case 7}
shows causality test case 7, for which the \co{r1 == r2 == r3 == 1}
result is to be allowed.
The \co{herd7} tool finds this result.

@@

This test case has everything
to do with reordering and nothing to do with OOTA cycles.
The \co{r1 == r2 == r3 == 1} outcome happens when lines~17, 10, 11,
15, 16 and~9 execute in that order.
Note that this execution order respects the semantic dependencies in
lines~10--11 and~15--16, and thus is not ruled out by hardware
dependency ordering.
It could have been ruled out if some of the accesses were acquire or
release, but here they are all relaxed.

@@

Listing~\ref{lst:Causality Test Case 8}
shows causality test case 8, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not report this result, but it does if
\co{r2} is initialized to one before being set to
\co{1 + r1 * r1 - r1}.

@@

Omniscient compilers might note that the computation on line~9
will have the value one when the load from \co{x} returns either
zero or one, which are the only values \co{x} and \co{y} can take.
Such a compiler could then realize that line~9 always sets \co{r2} to one,
independent of the value loaded in line~8.
Then there would be no semantic dependency between \co{P0()}'s load
and store, and thus no OOTA cycle relative to this compiler.

A less-capable compiler would leave the dependency in place,
and the resulting hardware dependency ordering would prevent
the OOTA cycle from being realized.

@@

Listing~\ref{lst:Causality Test Case 9}
shows causality test case 9, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not report this result.

@@

This is the same as causality test case~8 but with \co{P2()} added;
the store to \co{x} in \co{P2} can interfere with the formation of the
OOTA cycle.

Unlike in causality test case~8, the computation on line~9 must be
carried out because now the load on line~8 might return two.
Consequently \co{P0} and~\co{P1} both have semantic dependencies,
and hardware dependency ordering will prevent the OOTA cycle from
being realized.

@@

Listing~\ref{lst:Causality Test Case 9a}
shows causality test case 9a, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not report this result.

@@

This is the same as causality test case~9 except that here \co{P2()}
sets \co{x} to zero instead of two.
As as result the computation on line~9 is once again independent of
the value loaded on line~8, which makes the analysis nearly the same
as that of causality test case~8.

@@

Listing~\ref{lst:Causality Test Case 10}
shows causality test case 10, for which the \co{r1 == r2 == 1} and \co{r3 == 0}
result is to be forbidden.
The \co{herd7} tool reports this result.

@@

\co{P0()} and \co{P1()} have straightforward
semantic dependencies between their respective loads and stores, and
so hardware dependency ordering will prevent an OOTA cycle from being
realized.
\co{P2()} has no semantic dependency, and \co{P3()}'s store does not
occur in this execution because \co{r3} is zero.
Thus neither of them contributes to an OOTA cycle.

@@

Listing~\ref{lst:Causality Test Case 11}
shows causality test case 11, for which the \co{r1 == r2 == r3 == r4 == 1}
result is to be allowed.
The \co{herd7} tool reports this result.

@@

Like causality test case~7, this test case has nothing to do with OOTA cycles.
The result may be obtained by executing lines~20, 12, 13, 18, 19, 10, 11,
and~17 in that order.

@@

Causality Test Case 12 in
Listing~\ref{lst:Causality Test Case 12}
uses arrays, which are not yet supported by the \co{herd7} tool.

@@

An omniscient compiler could note that only \co{P0()} accesses
array \co{a[]} and act as if the code was simpler, but there would
still be semantic dependencies between \co{P0()}'s and \co{P1()}s
relaxed atomic loads and stores involving \co{x} and \co{y}.
The resulting hardware dependency ordering will prevent an OOTA cycle
from being realized, whether the compiler is omniscient or not.

@@

Listing~\ref{lst:Causality Test Case 13}
shows causality test case 13, for which the \co{r1 == r2 == 1}
result is to be forbidden.
The \co{herd7} tool reports this result.

@@

This is the same as causality test case~10 without the confounding
influence of \co{P2()} and~\co{P3()}.

@@

Listing~\ref{lst:Causality Test Case 14}
shows causality test case 14, for which the \co{r1 == r3 == 1} and \co{r2 == 0}
result is to be forbidden.
The \co{herd7} tool reports this result.

@@

In this execution, \co{P0()} has a semantic dependency from line~9 to
line~13 and \co{P1()} has a semantic dependency from line~18 to
line~21.
(There's also a semantic dependency from line~17 to line~21 but it
doesn't enter into this OOTA cycle, which involves only \co{x} and~\co{y}.)
The \co{memory_order_seq_cst} specification on line~17 has no important effect;
hardware dependency ordering will prevent the cycle from being realized.

@@

Listing~\ref{lst:Causality Test Case 15}
shows causality test case 15, for which the \co{r0 == r1 == r3 == 1}
and \co{r2 == 0}
result is to be forbidden.
The \co{herd7} tool reports this result.

@@

This execution has an OOTA cycle involving \co{u} and~\co{v}.
The dependency in \co{P0()} from line~13 to line~19 and the dependency
in \co{P1()} from line~24 to line~27 are both semantic.
The resulting hardware dependency ordering prevents the cycle from
being realized.

Having no load, \co{P2()} cannot participate in an OOTA cycle.

@@

Listing~\ref{lst:Causality Test Case 16}
shows causality test case 16, for which the \co{r1 == 2} and \co{r2 == 1}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be
expected because it would violate the C++ memory model's read-write
coherence rule regarding the modification order of an atomic object
(6.9.2.2p17 \co{[intro.races]}).
Because no abstract execution can obtain \co{r1 == 2} and \co{r2 == 1}, no
valid realization will either.

@@

Listing~\ref{lst:Causality Test Case 17}
shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, although it does
report a different OOTA cycle involving lines 11, 12, 16, and~17
(the \co{S17} value on line~3 of the \co{herd7} output).

@@

A compiler that treats atomics as quasi volatile could reason as follows:
Suppose that \co{P0()}'s load from \co{x} on line~8 returns the value 42.
In that case, the second load on line~11 can be merged with the first,
so that it also returns the value 42.
On the other hand, if the load on line~8 returns some other value then
line~10 will store the value 42 to \co{x}, and then the load on line~11
can be omitted in favor of using the value 42 stored by line~10.

Either way, the value of \co{r1} will be 42.
Therefore the loads from \co{x} cannot affect the store on line~17,
so there is no semantic dependency.
This allows the cycle to be realized,
but because the dependency in \co{P0()} is not semantic,
the cycle is not OOTA.

@@

Listing~\ref{lst:Causality Test Case 18}
shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result.

@@

This is the same as causality test case~17, even to the alternate OOTA
cycle reported by the \co{herd7} tool, except that line~9 tests
\co{r3} for equality to zero rather than non-equality to 42.

Perhaps surprisingly, this makes a difference.

@@

Listing~\ref{lst:Causality Test Case 19}
shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result but it does report the
OOTA cycle in line~3 of its output.

@@

This is nearly the same as the Simple OOTA cycle and causality test
case~4, with the addition of \co{P2()}.
Just as in those examples, the OOTA cycle cannot be realized by
compilers that treat atomic objects as volatile or that do
single-thread analysis and treat atomic objects as quasi volatile.

@@

Listing~\ref{lst:Causality Test Case 20}
shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result.

@@

This is the same as causality test case~19, even to the OOTA cycle
reported by the \co{herd7} tool, except that line~19 tests
\co{r3} for equality to zero rather than non-equality to 42.
The same analysis applies.
